---
title: "Response to Comments"
subtitle: "Individual Partisan Harm and Utility in Redistricting"
author: "Christopher T. Kenny (Work with Cory McCartan)"
date: "2021.11.05"
output: 
  beamer_presentation:
    slide_level: 2
    theme: "Singapore"
    colortheme: "beaver"
    toc: true
classoption: handout 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(here)
library(knitr)
```

# Abstract and Intro

## Abstract

Most existing standards for assessing partisan gerrymandering focus on the symmetric treatment of parties by electoral systems. We develop measures of utility and harm for individual voters by considering counterfactual redistricting plans, and use these notions to construct a normative standard of partisan fairness which is centered on the individual. We calculate these measures on real-world data by leveraging recent advances in redistricting sampling methodology, which allow researchers to generate large numbers of districting plans which reflect local legal and practical requirements. We apply our measures to congressional redistricting in New Jersey, and compare them to existing measures of partisan symmetry and competitiveness.

## Abstract

>- Start with "How do you measure a partisan gerrymander?" (Sayumi)
>- Use of "symmetric treatment of parties" is unclear (Shusei) 
>- Reverse s2 to (Adam)
>    -  "In contrast, we construct a new normative standard of partisan fairness which is centered on the individual." 
>- Not just different, highlight that it's better (Tyler)

## Intro

>- State contribution more prominently, currently paragraph 5ish (Adam, Jake, Tyler)
>    - Frame contribution more in measurement of harm in this context (Adam)
>    - Use intro to explain why individuals are needed for the legal side of harm
>        - Make clearer if we can bridge the legal-measurement gap (Avery)
>- Spend more time discussing other measures earlier (Shusei, Adam, Jialu, Adeline, Casey, Tyler, Soichiro)
>    - Define partisan [bias/symmetry/fairness] and efficiency gap (Sun Young, Shusei, Avery)
>- Spend more time reviewing the literature at large (Adam)
>- Add Section 2.3 to the Intro (Sooahn)

## Why care? What's the contribution? (Broadly: What's the frame?)

>- Why do we care about the *individual*? (Soubhik)
>- The focus on individuals isn't super novel, courtwise (Adam)
>- Make a starker contrast with the party-centric approaches
>- There's a literature on using statistics to guide ethical questions (Georgie)
>- When is this better than the existing? Is it ever worse? (Tyler)

# Notation, Utility, Harm, Assumptions

## Notation

>- _too_ much formality (Soubhik)
>- Subscripts vs. function inputs (Soubhik, Georgie)
>    -   $votes^(p)(d,v,q) \to V_{ijt}$
>- Delta H(q) with upper- and sub-script should be changed (Soichiro)
>- Add a glossary table (Soubhik)
>- Spend more time explaining the notation and its intuition (David, Shusei)
>- Only introduce the the notation (and figures) as needed (Georgie)
>- Specify the variables we take expectation with respect to (David, Sooahn)


## Utility

>- Do you need utility to get to harm? Is it periphary? (Jake)
>- Discussions of difference in local vs global utility seem interesting (Jake)
>- Discuss how these are related to each other (Jialu, Adeline, Soichiro)
>- Is strategic voting an issue for calculating utility? (Jialu)
>- Premise for utility is _Chris and Cory_ think it should be this way. Root it better in literature. (Avery)
>- Visualize this (Soichiro)

## Harm

>- Couldn't harm just be subpar utility? (David)
>- How is a voter harmed by not being rep'ed by their copartisan? (Jake)
>- Why use local measures and not global here? (Georgie)
>- What is the trade off of our harm measure compared to utilities alone (Avery)
>- Give it a cooler name (Tyler)
>    - Give it cooler notation (Soichiro)
>- Visualize this (Soichiro)

## Assumptions about votes/voters

>- Votes independent of districting plans seems too strong (Adeline)
>- Should we discuss turnout as a function of competitiveness? (Avery)
>- Too much reinventing the wheel re election modelling (Georgie)
>- How would votes being non-fixed impact measures? (Avery [and others])
>- The Individual may be unclear depending on how this is focused (Tyler)

# Mini Examples

## Examples

>- Mininois, Minichusetts, and Minisouri seemed to be more helpful than silly
>- Elaborate more on their usage and choice of _these_ examples
>- Appear a bit too early/move them out of Section 1 (Adam)

## Alternative examples

>- Can we better classify gerrymanders and non-gerrymanders by using enacted districts? (Jake)
>- Mini examples, but for measures (Casey)
>- Study all fifty-states and demonstrate that this is convincing everywhere (Tyler)

# Sampling

## Sampling

>- 1,000 seems small for a sample size (Adam, Jake)
>- Why is the set of legally feasible plans the relevant counterfactual? (Georgie)
>- Explain "Superfluous partisan considerations" (Avery)

# New Jersey

## New Jersey

>- Explain the structural advantage of Democrats in New Jersey (Tyler)

```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders. Census-designated places are shaded in gray and county lines in white.", fig.width=6.5, fig.height=3.5}
include_graphics(here("paper/figures/nj_maps.pdf"))
```

## Figure 5

```{r nj-vars, fig.cap="Distributions of districting measures for simulated New Jersey distriting plans, the Republican gerrymander (red), and the Democratic gerrymander (blue).", fig.width=6.5, fig.height=3.25}
include_graphics(here("paper/figures/nj_vars.pdf"))
```

## Figure 5

>- What's the takeaway? (Soubhik)
>- Add names, not just symbols to subtitles (most people)
>    - Reminders of what they mean (Sooahn)
>- Is this validation or comparison?
>    - Yes!


## Figure 6

```{r nj-district-shares, fig.cap="Distribution of district vote shares, with districts ordrered by baseline Democratic two-part vote share. Displayed on top are the district vote shares for six example plans, each extremal for some combination of districting measures.", fig.width=5.8, fig.height=3.5}
include_graphics(here("paper/figures/nj_district_shares.pdf"))
```

## Figure 6

>- Fairly confusing figure, please elaborate (Sun Young, Jialu, Avery)
>    - Everyone who mentioned it was for confusion, which is a strong signal
>- Spend more time _explaining_ the figure, rather than _describing_ the figure

## Section 5 (Connections to Partisan Symmetry and Fairness Standards)

>- Incorporate uncertainty calculation into the S5 metric calculations (David)
>- Make this section smaller and instead use this to motivate the discussion (Sayumi)


# Going Forward

## Formatting

>- Add a longer discussion section that better frames the contribution (Jake, Sayumi)
>- Many, many sections. What's the target journal? (Sun Young)
>- Move propositions to appendix (Georgie)
>- Utility (or maybe global utility) to appendix (Tyler)
>- Figure 3 is hard to digest with two color scales (Sayumi, Casey)

## Expanding Focus 

>- How does this extend to multiple groups? (Adeline)
>- Handwaving towards multiple parties might be too little for potential strategic considerations (Sooahn)

