---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed. 
      The authors thank participants of the Harvard American Politics Research Workshop for helpful comments and discussion.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Abstract here.
keywords:
  - representation
  - redistricting simulation
  - individual standing
bibliography: references.bib
biblio-style: apalike
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: sentence
---

```{r SETUP, include=FALSE}
library(knitr)
library(here)

opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
               fig.path=here("paper/figures/"), fig.align="center",
               fig.width=(8.5-2*1)/2, out.width="100%", fig.asp=0.8)
```

# Introduction

-   what makes a good redistricting plan?

-   on the partisan side, lots of attention on the fairness to parties

    -   partisan symmetry [@katz2020]

    -   seats-votes curve

-   on the racial side, ... ?
    (VRA focus? looking at race of candidates vs choice? what? I don't know)

-   courts, on the other hand, have focused on individual standing [@1964reynolds; @1962baker]

-   **we argue that individual standing/harm is an excellent lens to unify the study and measurement of redistricting plans**

-   core idea: counterfactuals + voter choice.
    Is a voter more likely to be represented by the candidate of their choice under a different redistricting plan?
    If so, they've been harmed.

    -   modern simulation methods allow us to produce realistic counterfactuals

    -   by analyzing how harms fall differentially along party, race, ideological, etc. lines, we can connect individual harm to more global measures based on parties, etc.

## Outline of the Paper

Section \ref{sec:motivation} motivates our framework for measuring harm and illustrates the key concepts in the context of partisan gerrymandering for three simple examples.
Section \ref{sec:indiv-harm} formalizes the intuition of Section \ref{sec:motivation} and defines a general measure of individual and differential harm.
Section \ref{sec:estimating} discusses estimating these measures from observed election results.
Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
Section \ref{sec:empirical} applies the proposed measures to congressional representation in New Jersey and primary elections for the New York City Council.
Section \ref{sec:discussion} concludes.

# Motivation {#sec:motivation}

To illustrate the concepts and measures we introduce, we will make use of three running examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
Each example state has the same population and should be divided into 3 districts with population between 54 and 66.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
Schematics for these examples are shown in Figure \ref{fig:mini-schematic}(a)--(c).

```{r mini-schematic, fig.cap="Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts. The total voting population for each example is 180. White lines delineate precincts."}
include_graphics(here("paper/figures/minis_schematic.pdf"))
```

## Minichusetts: counterfactuals matter

-   40% GOP but no GOP districts.
    Is this unfair?

-   E-gap would say so.
    Partisan symmetry: ?.

-   But impossible to draw a GOP district

-   Without a remedy (counterfactual), there can be no harm

## Mininois: individual harm

-   show a fair map and a gerrymandered map

-   zoom into 1 or 2 precincts & calculate individual harm of D & R voters in those precincts

-   show map of differential harm (with

## Minissouri: differential harm

-   show a fair map and a gerrymandered map

-   tally harm by party

-   end with transition: there are natural avenues for generalization

```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm)."}
include_graphics(here("paper/figures/mini_harm.pdf"))
```

Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure \ref{fig:mini-util-opt}.
As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).

# Individual Representational Harm {#sec:indiv-harm}

-   this section: formalize and generalize the concepts in Section \ref{sec:motivation}.

## Setup

-   Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.

-   Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
    In an election, voters either vote for a candidate running in their district, or abstain from voting.
    We write $v_i$ [**CM: should this be** $v(i)$?]
    to mean the candidate that individual $i$ votes for in election $v$.

-   Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q_i$ [**CM:** $q(i)$?]
    indicating the district that individual $i$ lives in.
    The set of all possible districting plans is $\Plans$.

-   For an election $v$, we can tally the votes for each candidate; the candidate with the most votes wins.
    We denote the winner in district $d$ of districting plan $q$ as $\win_d(v,q)$.

-   explain that $v$ and $q$ are necessarily related: a different districting plan may mean different candidates run.

## Harm

To be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
Then we may define the harm to an individual $i$ as $$
h_i(v,q,\tilde q) = \ind\{v_i \neq \win_{q(i)}(v, q)\}
        \ind\{v_i = \win_{\tilde q(i)}(v, \tilde q)\},
$$that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.

## Multiple Elections or Districting Plans

So far, harm is defined for a single election outcome $v\in \Elecs$ and districting plan $q\in\Plans$.
But realistically we will need to consider harm averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans, or even over a set of individuals with some shared characteristic.

To allow this, we replace the fixed election and districting plans with random variables $V\in\Elecs$ and $Q,\tilde Q\in\Plans$.
The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V$, $Q$, and $\tilde Q$ are mutually independent, though this is not required for our theoretical analysis.
We will also consider an individual $I$ taken uniformly at random from $\Voters$, with $I\indep V,Q,\tilde Q$.

Then we may define the harm for a (randomly selected) individual as $H \dfeq h_I(V, Q, \tilde Q)$.
Thus, for example, the harm for a single individual $i$ becomes $H\mid I=i$; and the electorate's expected harm under districting plan $q$ becomes$$
\E[H\mid Q=q] = \Pr[V(I)\neq\win_{q(I)}(V, q)
    \text{ and }V(I)=\win_{\tilde Q(I)}(V, \tilde Q)],
$$that is, the probability that a randomly selected voter is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.

**--------------- CM: this is as far as I've gotten. ---------------**

# Differential Harm and Fairness {#subsec:diff-harm}

Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
This is easily expressed using our notation above; $Q$ is *fair* in comparison to counterfactual $\tilde Q$ if $\E[H\mid V(I) = p]$ is constant across parties $p\in\Parties$.

To quantify deviations from fairness, we introduce of a measure of \textit{differential harm}, defined as $$
  \Delta^{p_1}_{p_2}H(q) \dfeq \E[H\mid V(I)=p_1, Q=q] - \E[H\mid V(I)=p_2, Q=q],
$$ for two parties $p_1$ and $p_2$.
So $\Delta^{p_1}_{p_2}H(q)=0$ for all $p_1,p_2\in\Parties$ if and only if $q$ is fair.
From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta^\dem_\rep H(q)$.
But the same concepts and definitions naturally carry over to the multiparty setting.

Despite being derived from a local harm measure, $\Delta H(Q)$ is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
Yet unlike a simple measure of seats, $\Delta H(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to partisan fairness.
In other words, $\Delta H(Q)$ provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
This connection further supports the use of total seats as a summary statistic in redistricting simulation methods, as advocated by @deford2020symmetry.

```{=tex}
\begin{restatable}{prop}{propfairness} \label{prop:fairness}
If turnout is equal across districts, i.e., if $\sum_{p\in\Parties} \votes^{(p)}(d,v,q)$ is constant across $d$, $v$, and $q$, then for a specific districting plan $q\in\Plans$ and political party $p\in\Parties$, we have
\begin{align*}
    \Delta H(q) = \E[\seats(p, V, Q)\mid Q=q] - \E[\seats(p, V, \tilde Q)].
\end{align*}

\end{restatable}
```
# Estimating the Harm Measures {#sec:estimating}

The measures $U^{(G)}, U^{(L)}, H$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plans, $\pi$.
Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
Either way, past election data and an enacted districting plan alone are not enough to calculate these measures.

While not strictly necessary, modeling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
Practically, this translates to the assumption that individual voters would vote for the same political party no matter the districting plan.
To the extent that voters reward or punish an incumbent party for the districting plan they adopt, or that an adopted districting plan affects which candidates run in a district, this assumption is violated.
But as our suggested model for election outcomes will not account for candidate factors or policy preferences, the marginal effect on our conclusions will be minimal.

## Modeling Districting Plans

It is difficult to construct a tractable probability distribution for districting plans, since the space of plans is discrete and lacks a natural metric structure.
Thankfully, in recent years a number of algorithms have been developed which sample districting plans from specific distributions [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These distributions can generally be characterized as maximum-entropy distributions on the space of all valid districting plans, in combination with several moment constraints.

For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans which are geographically contiguous, have district populations within a deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H$ and $\Delta H(q)$.

## Modeling Election Outcomes

Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
As far as estimating measures like $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
In the interest of fairness, the model should not make any biasing assumptions about the future (e.g., that demographic shifts will favor one party, or that past trends in election returns will continue in the same direction).

We propose a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
Each election outcome over the period of interest (generally ten years or less) is modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
The baseline itself can be established by averaging the results of all available statewide elections over the past decade or other suitable period.

The proportionate uniform shift model is not mathematically equivalent to the uniform partisan swing model of @butler1951swing, but is near-identical for most precincts and districts. Despite their simplicity, uniform-swing--type models are highly accurate for out-of-sample prediction [@katz2020], and are intuitive and easy to implement.

Proportionate shift, where we assume that a constant fraction of *voters of one party* in each precinct switch to the other party, has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
For measures like $H$ and $F$ which consider voters in each party separately, this helpfully simplifies computations.

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlikea generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^1]
Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
\gamma(s_k) = \gamma(\overline s_k) + \eps, \qquad \eps \sim \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

[^1]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.

More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
For example, one could model district deviations as a systematic national shift plus a local idiosyncratic effect, as is done in the Stochastic Uniform Partisan Swing model of @gelman1994unified.
For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
Appendix \ref{app:formulas} contains the specific formulas used to compute $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.

# Connections to Partisan Symmetry and Fairness Standards {#sec:connections}

In the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders. Specifically, these measures aim to quantify deviations from partisan symmetry, fairness, or competitiveness. Some of the most commonly used are perhaps the *efficiency gap* [@stephanopoulos2015], *partisan bias* [@tufte1973pb; @king1987democratic], and the *mean-median* measure [@mcdonald2015unfair].
All of these measures may be estimated using district vote shares alone, and as we discuss next, are to varying degrees related to our proposed measures.

## The Efficiency Gap

The efficiency gap ($EG$) interrogates how a party's votes are wasted under a districting plan.
By subtracting the wasted votes for Democrats from those for Republicans, and normalizing by the total votes cast, it offers a signed measure that is simple to interpret: $EG=0.05$, for example, indicates that 5% more Democratic votes are wasted than Republican ones.
The general recommendation is that an efficiency gap larger than two Congressional seats (i.e., 2 divided by the number of districts), or 0.08 for state legislative plans, indicates a partisan gerrymander.

A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
These concerns to not apply to our proposed measures.

First, $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ are all averaged over possible election outcomes, and so do not suffer from discreteness.
And as discussed above, $U^{(G)}$ and $U^{(L)}$ both reward proportionality.
Of course, proportionality is not always possible, given a state's political geography.
The harm measures $H$ and $\Delta H(q)$ both reflect this by comparing to a set of counterfactual districting plans.
This addresses one inherent shortcoming of the efficiency gap, illustrated best by the state of Massachusetts.
According to the efficiency gap, Massachussetts' congressional districting plan is a Democratic gerrymander ($EG=-0.14$).
But it is impossible to draw a single Republican congressional district in the state.
Consequently $H$ and $\Delta H(q)$ are everywhere 0.

Finally, the fact that the efficiency gap is minimized for districts with 75%--25% vote shares is a direct consequence of comparing the raw number of wasted votes, rather than the share within each party.
In contrast, $\Delta H(q)$ is calculated by comparing harm averaged first within each party, avoiding this deficiency.

As we will see in the application below, often there is substantial correlation between the efficiency gap and $\Delta H(q)$.
Yet in more unusual or tricky cases, the efficiency gap may point to a partisan gerrymander while $\Delta H(q)$ indicates a fair outcome or an unfair advantage for the opposite party.

## Partisan Bias

Partisan bias is probably the most widely-used measure of partisan symmetry.
The bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.
Each value of $s$ yields information about a different point on the seats-votes curve.

In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practictioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.

Fundamentally, however, $\beta(s)$ measures partisan symmetry, and not partisan harm or utility.
It should not be surprising, then, that it can and does "disagree" about the partisan effects of a particular plan.
These "disagreements" are really just differences in the target quantity of interest.

```{=html}
<!--
1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
2.  Focuses on only one part of S-V curve, which may not be relevant

Notably, as the quantity of interest is not partisan symmetry, our measures do not require the existence of a seats-votes curve. This is consequential for the study of congressional redistricting. As a seats-votes curve is defined for a legislature, rather than a delegation, to understand the impact of single state's congressional map, you would be forced to consider how every other state drew their maps [@katz:king:rose:21a]. Thus, if this taken as the definition of a seats-votes curve, then measures of partisan symmetry lack diagnostic power to identify gerrymanders for the delegations that academics, legal scholars, and the general public are often focused on. The individualized focus of our measures are, however, portable to congressional, state legislative, or any other form of districting and elections.
-->
```
## Mean-Median Difference

The mean-median difference straightforwardly measures the difference between the mean and median district vote shares.
In practice, the mean-median differences is moderately correlated with other partisan redistricting measures.
However, considerations of the mean and median district are only related to individuals insofar as they describe the general propensity to be harmed or benefited by a particular map.
As [@katz2020] discuss, the mean-median difference can be interpreted as an estimator of *votes-denominated bias*, another value derived from the seats-votes curve.

## Districting Plan Competitiveness {#sec:competitive}

The notion of district *competitiveness*, while related to the slope of the seats-vote curve around $\half$, is not the same as partisan symmetry.
However, one commonly used measure of competitiveness, from @tamcho2016, is related to our local utility measure $U^{(L)}$.
The measure of @tamcho2016, which we will denote by $\mathrm{comp}_\alpha(v,q)$, contains two terms: one, $T_p(v,q)$, measuring the closeness of district vote shares to 50%, and the other, $T_e(v,q)$, measuring the closeness of the overall share of seats to 50%; the relative importance of these two terms is controlled by $\alpha$.
It is this first term which his closely related to $U^{(L)}$, as stated in Proposition \ref{prop:comp} below.

```{=tex}
\begin{restatable}{prop}{propcomp} \label{prop:comp}
If turnout is equal across districts, for an election $v\in\Elecs$ and districting plan $q\in\Plans$ and two political parties, we have
\begin{align*}
    \E[U^{(L)}\mid V=v, Q=q] &= \frac{1}{2|D|}\sum_{d\in D}
        \qty|\frac{\votes^{(p_a)}}{\votes^{(p_b)} + \votes^{(p_b)}}-\half| \\
    &= \half T_p(v,q) = \frac{\mathrm{comp}_\alpha(v,q)}{1+\alpha T_e(v_q)}.
\end{align*}

\end{restatable}
```
Importantly, $\mathrm{comp}_\alpha(v,q)$ is designed to be *minimized*---smaller values correspond to more competitive districts.
But, as briefly discussed above, this actually runs counter to the local utility measure, which encourages lopsided districts with as many similarly-minded voters as possible.
@brunell2010 discusses this point in much more detail, arguing that competitive districts should be avoided where possible.
Buttressing our choice of local utility function, he analyzes ANES survey data to show that voters approve more highly of their representative and of Congress overall when their representative votes in alignment with their preferences.

# Application: New Jersey {#sec:empirical}

To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
As discussed above, many gerrymandering measures function well in highly competitive states, but may produce more counterintuitive results in other states.
New Jersey, with an average Democratic two-party vote share of 56.6% across our two baseline elections (2016 presidential and 2018 senatorial), will demonstrate that this is not a problem unique to lopsided states.
Additionally, New Jersey has 12 congressional districts, which is enough to produce some variability in seats outcomes across counterfactual plans, but not so many as to render trivial the challenges of discreteness.

```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders. Census-designated places are shaded in gray and county lines in white.", fig.width=6.5, fig.height=3.5}
include_graphics(here("paper/figures/nj_maps.pdf"))
```

Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with two hypothetical gerrymanders of the state, which were generated using the short-burst optimization approach of @cannon2020.
The Democratic gerrymander has 11 Democratic seats and 1 Republican seat at baseline, while the Republican gerrymander has 6 Democratic and 6 Republican seats.

We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^2]
The distribution of these measures is shown in Figure \ref{fig:nj-vars}.

[^2]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].

The efficiency gap and $\Delta H(q)$ both correctly classify the gerrymandered plans as such, and with the correct sign.
Both gerrymanders also have a higher average harm than any of the simulated plans.
In contrast, under the partisan bias measure the Republican gerrymander is completely unbiased, and according to the mean-median measure, both gerrymanders strongly favor Democrats.

For the utility measures, the Democratic plan is a low-utility outlier.
But the Republican plan is near the median for both global and local utility.
How to square this with the finding that the Republican plan is an unfair gerrymander which harms more voters than expected under a neutral process?
As the distribution of expected Democratic seats shows, Democrats are structurally favored in New Jersey, able to win a larger fraction of seats than their statewide vote share under a neutral process.
Plans which actively counteract this to try to draw proportional districts, like the Republican gerrymander, will necessarily harm Democratic voters versus the neutral counterfactual.
But in doing so they will create a more proportional plan that will tend to pack voters of each party into more homogeneous districts, increasing both global and local utility.

```{r nj-district-shares, fig.cap="Distribution of district vote shares, with districts ordrered by baseline Democratic two-part vote share. Displayed on top are the district vote shares for six example plans, each extremal for some combination of districting measures.", fig.width=6.5, fig.height=3.9}
include_graphics(here("paper/figures/nj_district_shares.pdf"))
```

This is manifested in the district-level vote shares shown in Figure \ref{fig:nj-district-shares}.
Within each districting plan, we number districts in order of their Democratic vote share, with district 1 the least Democratic.
The figure shows these vote shares for six selected districting plans, and compares them to the distribution of shares for the entire sample.
The selected plans themselves are shown in Appendix \ref{app:nj}.

The fairest, best overall utility, and high global, low local utility plans have 8 Democratic and 4 Republican seats at baseline, but the best-utility plan has more extreme vote shares, i.e., less competitive districts, while the high global, low local utility plan has vote shares closer to 50%, with the fairest plan falling somewhere in the middle.

In contrast, the low global, high local utility plan has 9 Democratic seats, 1 Republican seat, and 2 highly competitive seats at baseline.
While generally local utility encourages proportionality and noncompetitive districts, conditional on low global utility (poor proportionality) the local utility measure encourages a Democratic gerrymander which satisfies the state's majority of Democratic voters, at the expense of the minority of Republican voters.

The fair but low local utility plan has a similar vote share pattern to the fair plan, but overall with more competitive districts, which are generally penalized by the local utility measure.
Conversely, the unfair but high local utility plan has 9 Democratic seats, 2 Republican seats, and 1 highly competitive seat at baseline---an unfair Democratic gerrymander---while keeping district vote shares relatively noncompetitive.

# Discussion {#sec:discussion}

This paper is an attempt to bridge the quantitative gap between votes and representation.
We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.
This focus naturally leads to a new normative standard of partisan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters of each party.
This may occur regardless of whether the districting plan achieves partisan symmetry.

This approach offers new purchase into the foundational questions of how to quantify the harms and benefits that accrue to voters in district-based democracies, where the rules of defining districts are fixed, but the choice of how to how to draw the districts remains a policy question.
While the overall framework is flexible, we offer a selection of simple yet interesting measures for improving the individual partisan implications of a districting plan.
Further work is needed to characterize how existing redistricting requirements and practices, such as the number or size of districts, or encouragement of competitive districts, interact with these measures.

# References {.unnumbered}

::: {#refs}
:::

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}
```
# Proofs of Propositions {#app:proofs}

```{=tex}
\propfairness*

\begin{proof}
\end{proof}

\propcomp*

\begin{proof}
\end{proof}
```
# Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}

```{=html}
<!--

1.  Global utility is discrete, so we can compute the range of shifts that produce each number of seats, and the average vote share for those ranges.
    Combine these

2.  To be computationally efficient, local utility requires linearizing the vote share curve.

3.  For voter fairness, PUS means that relative share of party's voters in each precinct remains the same.
    Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)

Formulas here.
-->
```
# Districting Measures in New Jersey {#app:nj}

```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures. Color indicates the number of Democratic seats won (ranging from 7 to 10). Below the diagonal, numbers indicate the absolute value of the correlation between variables.", fig.width=7, fig.height=7}
include_graphics(here("paper/figures/nj_pairs.pdf"))
```

```{r nj-ex-maps, fig.cap="Example districting plans selected according to their scores on $\\Delta H(q)$, $U^{(G)}$, and $U^{(L)}$. District shares for these example plans are shown in Figure \\ref{fig:nj-district-shares}.", fig.width=7, fig.height=7}
include_graphics(here("paper/figures/nj_ex_maps.pdf"))
```

# Misrepresentation Map

The intuition that drives these measures is the misrepresentation map.
Consider a state that is partitioned into smaller geographies, typically called precincts or voter tabulation districts.
These units act as the atoms of redistricting.
We assume that they are not split in the redistricting process.
Thus, districts are made of whole precincts.
For a given precinct, we can thus calculate the expected party share of the district that a precinct is in for any given plan.
The expected party share can take many forms, as mentioned in *Modeling Election Outcomes*.
By combining the information about the geographic precinct, we can measure the expected party share that an individual living in that precinct would expect to experience.

Visually, we can consider a map colored by party share of the district that each precinct is assigned to.
Most often, we represent this in a blue to red gradient which offers the typical Democratic or Republican representative dynamic.
Given a counterfactual, we can then take the difference of the expected party share, precinct-by-precinct.
This offers a precinct-level coloration based on difference in expected party share at the district level.
In the typical two-party scenario, using Democratic two-party vote share, a positive difference would indicate that the individuals in those precincts are likely to be represented by Democrats.
A negative difference would indicate that the individuals in those precincts are likely to be represented by Republicans.
Thus, positives would be map-matically represented by blues and negatives by reds.

Note that this takes a similar visual form as a dislocation map, though these are conceptually distinct [@defoeubarodd21].
The dislocation map indicates within a single plan how nearby individuals are likely or unlikely to be represented by the same party.

We focus largely on the case of a misrepresentation map that is the difference between a single enacted districting plan and an ensemble of alternate, legal districting plans.
This does not need to be the case, however.
A misrepresentation map can indicate the difference between two singular maps, calculating the difference in expected district-level party representation.
This would allow direct computation of utility, fairness, or other metrics based on the difference in the two maps.
Similarly, given two sets of ensembles, likely generated under different target distributions, we can create a similar misrepresentation map that would allow us to estimate interesting measures, such as the differential harm caused by the inclusion of an additional criteria in state district drawing rules.
