---
title: Measuring Individual Partisan Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: Footnote stuff
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - partisan fairness
  - representation
  - redistricting simulation
  - individual standing
bibliography: references.bib
biblio-style: apalike
output: 
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
---

```{r SETUP, include=FALSE}
library(knitr)
library(here)

opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
               fig.path=here("paper/figures/"), fig.align="center",
               fig.width=(8.5-2*1)/2, out.width="100%", fig.asp=0.8)
```

# Introduction

What is the question?

Districting is a complicated process.
Residents of a state or locality must be partitioned into districts.
These districts must adhere to many constraints, often requiring components such as contiguity and equal population.
In drawing the districts, it's easy to lose sight of the individuals impacted by district drawing.
Often, we consider how parties succeed or fail under the electoral environments generated by enacted plans.
Yet this misses that parties are formed by voters who ally with parties.
Rather than considering parties, we ask how individuals are harmed or benefited by the districting process?

Why does it matter?

Why is it hard?
Decades of

What did we do?

What did we learn?

What is the paper's contribution?

```{=html}
<!--
Understanding fairness in electoral systems has long suffered from a focus on parties over the individuals who vote for them.
There is no doubt that parties are interesting units in electoral discussions.
They offer stability over candidates, information for the under-informed, and shortcuts for understanding preferred vote choices.
Yet, court histories and academic traditions value the harm of the individual over the harm to a transient political body, such as the party.

In this article, we offer an approach to understanding partisan fairness through a lens of individual harm and utility.
This does not supplant existing understandings of partisan fairness, but instead offers an alternative rooted in understanding how the individual potential voters are impacted by districting plans.

Several existing methods of redistricting simulation rely on randomness, such as Random Seed and Grow (Chen and Rodden 2013) or Recombination (Deford et al. 2021).
Moving beyond these heuristic methods, we can use sampling methods that have defined target distributions to ensure that we characterize the space of applicable plans properly.
Methods which operationalize a statistics-based approach include Fifield et al. 2020, McCartan and Imai (2020), and Autrey et al (2020).
Using these methods advance us beyond the critiques of Katz, King, and Roseblatt (2020) by sampling plans using weights within a Gibbs distribution, rather than assuming that uniform sampling weights alone are sufficient.

Metrics often associated with partisan symmetry, such as TODO, are indicative of violations of partisan symmetry.
Notably, these are only defined for full legislatures (see Katz, King, and Rosenblatt (2021)).
As such, theory-based interpretations of these measures do not indicate signs of unfairness to parities in the congressional district setting, unless they are consistent across many, many states.
We offer measures of partisan harm which are rooted directly in the representational harms inflicted upon individuals by those who choose to draw a map.

These measures are deeply connected to the law and constitutional understandings of partisan fairness, due to their related roots in the individual.

-   introduce partisan symmetry

    -   Katz, King, Rosenblatt 2019; Deford et al 2021; Katz, King, Rosenblatt 2021

-   Election law side is very indivudalized

    -   Standing discussions

        -   Lujan v Defenders of Wildlife (1992)
        -   Shaw v. Reno (1993) -- racial gerrymandering
        -   Issacharoff and Karlan 1998

-   Recent, related work

    -   EGap
    -   Partisan Dislocation

-   Important to consider benefits and harms to individual voters

-   Power of Ensemble Analysis

    -   Recent advances allow us to make claims about harm based in counterfactuals that are realistic and statistically useful
-->
```
## Running Examples

To illustrate the notation and measures we introduce, this article will make use of three running examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
Each example state has the same population and should be divided into 3 districts with population between 54 and 66.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
Schematics for these examples are shown in Figure \ref{fig:mini-schematic}(a)--(c).

```{r mini-schematic, fig.cap="Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts. The total voting population for each example is 180."}
include_graphics(here("paper/figures/minis_schematic.pdf"))
```

## Outline of the Paper

Section \ref{sec:voter-util} provides the necessary notation and introduces our measures of voter utility.
Section \ref{sec:indiv-harm} builds a measure of individual harm and a derived measure of partisan fairness from this notion of utility.
Section \ref{sec:estimating} discusses estimating these measures from observed election results.
Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
Section \ref{sec:empirical} applies the proposed measures to a real-world example to illustrate their value and these comparsions.
Section \ref{sec:discussion} concludes.

# Voter Utility in Election Outcomes {#sec:voter-util}

The utility voters gain from a particular election outcome naturally varies across the electorate, depending on factors such as political engagement, partisanship, race and ethnicity, individual policy preferences, and individual candidate attributes.
Here, we restrict our attention to what we believe are the two dominant contributors to partisan utility: which party represents a voter's district, and how many overall seats a voter's preferred party wins.

## Notation and Setup

Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate from one of a set of political parties $\Parties$ to represent each of a set of legislative districts, indexed by $\Distrs$.
We take these three sets to be fixed.

Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
In an election, voters either vote for a candidate running in their district, or abstain from voting.
So we write $v:\Voters\to\Parties\cup\{\textsc{abstain}\}$, with $v(i)$ the party that voter $i\in\Voters$ voted for in election $v$.

Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is formally represented by a function $q:\Voters\surjto \Distrs$.
The set of all possible districting plans is $\Plans$.[^1]

[^1]: Note that these districting plans do not depend on the election; as a result, we implicitly assume that voters do not move residences during the period under consideration.
    This is not a necessary restriction and can be dealt with formally by first mapping voters to precincts (varying by election) and then mapping precincts to districts (constant across elections).
    Once we aggregate or average across all voters in a party or district, the differences between these two approaches disappear.

Votes may be tallied at the district level, by party, by defining for each party $p\in\Parties$ $$
    \votes^{(p)}: \Distrs\times\Elecs\times\Plans \to \N, \qquad 
    \votes^{(p)}(d, v, q) = \sum_{i\in q^{-1}(d)} \ind_{\{p\}}(v(i)).
$$ The political party of the candidate with the most votes wins each district, as captured by the function $$
    \win: \Distrs\times\Elecs\times\Plans\to P, \qquad 
    \win(d, v, q) = \argmax_{p\in\Parties} \votes^{(p)}(d, v, q).
$$Thus an individual voter $i\in\Voters$ is represented after election $v$ by political party $\win(q(i), v, q)$, and each political party can be said to win a number of seats, formalized as$$
    \seats:\Parties\times\Elecs\times\Plans\to\N, 
    \qquad \seats(p,v,q) = \sum_{d\in\Distrs} \ind_{\{p\}}(\win(d,v,q)).
$$

## Global Outcome Utility

We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
    u^{(G)}:\Voters\times\Elecs\times\Plans\to\R, 
    \qquad u^{(G)}(i,v,q) = g\qty(\frac{\seats(v(i), v, q)}{|\Distrs|}),
$$for $g:[0,1]\bijto\R$ an increasing function.

What kind of electoral systems does this utility function encourage, if the goal is maximizing total voter utility (or, equivalently, average voter utility)?
It depends on the form of $g$, as the following result describes.

```{=tex}
\begin{restatable}{prop}{propglobal} \label{prop:global}
Let $S=\{s_p:p\in\Parties\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the popular vote.
Define the total utility for this rule as \[
    u_{\mathrm{tot}}^{(G)}(f, v) \dfeq \sum_{i\in\Voters} g\qty(f\qty(s_{v(i)})).
\] 
Then if $g$ is convex, $u_{\mathrm{tot}}^{(G)}$ is maximized for $f(x)=\ind_{[\max S, 1]}(x)$, and if $g$ is strictly concave, the maximizer of $u_{\mathrm{tot}}^{(G)}$ will be a strictly increasing function.
In particular, if $g(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $u_{\mathrm{tot}}^{(G)}$.

\end{restatable}
```
Thus, if we are willing to adopt the logarithm for the functional specification of $u^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.

## Local Outcome Utility

The specification of the utility of the local election outcome is more straightforward.
We define$$
    u^{(L)}:\Voters\times\Elecs\times\Plans\to\R, 
    \qquad u^{(L)}(i,v,q) = \ind\{v(i)=\win(q(i), v, q)\},
$$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.

Attempting to maximize the average local utility across the electorate would lead to a districting plan where every voter is assigned to a district with other voters of the same party preference; in this way, every voter's utility would be $1$.
Whether such an districting plan is possible depends on the geographical layout of voters and the legal constraints on the districting process---must districts be contiguous, balanced in population, constructed from whole precincts, geographically compact, and so on.
But the general tendency of this utility measure (in the context of a single election, at least), like the global utility measure, will be to push plans towards proportionality.
Unlike with the global measure, a specific kind of proportionality is preferred: that which creates as many lopsided districts as possible (what is often termed a *bipartisan gerrymander*).
This point is discussed further in Section \ref{sec:competitive} below.

## Examples

To illustrate these two utility measures, Figure \ref{fig:mini-util-opt} shows three plans for our running examples which maximize both global (with $g=\log$) and local utility.[^2]
In Mininois and Minissouri, the overall division of seats is as close to proportional as possible.
In Minichusetts, however, it is impossible to give any seats to Republicans (as long as districts are constructed from precincts); all plans have the same utility.
In none of the three states is it possible to create a perfect bipartisan gerrymander, but the plans for Mininois and Minissouri get as close as possible by packing Democratic voters into cities (and Republican voters into rural areas, where feasible).

[^2]: These examples present a slight complication---what to do when a party wins no seats?
    The easiest solution, perhaps, is to define the utility to that party's voters to be $-1$; this will make the utility well-defined while still encouraging proportionality.

```{r mini-util-opt, fig.cap="Utility-maximizing redistricting plans for the three running examples, with each district shaded by its partisan lean."}
include_graphics(here("paper/figures/minis_util_opt.pdf"))
```

## Multiple Elections or Districting Plans

The utility measures above are defined for a single election outcome $v\in \Elecs$ and districting plan $q\in\Plans$.
But often we will want to consider utility averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans.

To allow this, we replace the fixed election districting plan with random variables $(V,Q)\sim \pi$, where $\pi$ is a joint probability measure on election outcomes and districting plans.
The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V\indep Q$, though this is not required for our theoretical analysis.
We will also consider a voter $I$ taken uniformly at random from $\Voters$, with $I\indep\pi$.
This has the effect of treating voters exchangeably and valuing their personal utilities equally.

Then we may define the utilities for a (randomly selected) voter as $$
    U^{(G)} \dfeq u^{(G)}(I, V, Q) \qand
    U^{(L)} \dfeq u^{(L)}(I, V, Q).
$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $v$ becomes $\E[U^{(L)}\mid V=v]$, and so on.

# Towards Individualized Representational Harm {#sec:indiv-harm}

Utility may be defined solely in terms of realized election outcomes and districting plans.
Harm, we argue, cannot.
Harm is not the same as disutility, for to be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
Put another way, in the context of districting plans, if no alternative districting plan could provide more utility to a voter, then that voter was not harmed by the districting plan chosen.

## Individual Harm

To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
Then we may define the harm to a voter as $h:\Voters\times\Elecs\times\Plans^2\to\R$, \begin{align*}
    h(i,v,q,\tilde q) &= (1 - u^{(L)}(i, v, q))\cdot u^{(L)}(i, v, \tilde q) \\
    &= \ind\{v(i)\neq\win(q(i), v, q)\}
        \ind\{v(i)=\win(\tilde q(i), v, \tilde q)\},
\end{align*}that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.

Harm is defined using the local, not global, utility measure, because inherently, global utility (and thus harm) will be constant for all voters of a particular party.
In fact, a harm measure defined analogously using global utility would essentially be a difference in seats won by a voter's party between plans $q$ and $\tilde q$ (the exact form would be affected by the function $g$).
This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis]. **[CM: revisit ¶ based on NJ findings]****.**

As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
Make the counterfactual plan a random variable $\tilde Q$, with $(V,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
Then the harm to a random voter is $H \dfeq h(I, V, Q, \tilde Q)$.

The expected harm to an individual $i\in\Voters$ voter then becomes $$
   \E[H\mid I=i] = \Pr[V(i)\neq\win(Q(i), V, Q)
    \text{ and }V(i)=\win(\tilde Q(i), V, \tilde Q)],
$$

that is, the probability that the voter is misrepresented under the adopted plan but would not have been under a counterfactual.

```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm)."}
include_graphics(here("paper/figures/mini_harm.pdf"))
```

Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure \ref{fig:mini-util-opt}.
As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

## Fairness

From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---since that will simplify the notation and exposition.
But the same concepts and definitions naturally carry over to the multiparty setting.

Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
This is easily expressed using our notation above; the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
    F = \E[H\mid V(I) = \dem] - \E[H\mid V(I) = \rep];
$$obviously the ordering of $\dem$ and $\rep$ is immaterial.
Informally, this quantity is the difference in the probability that a voter of each party is represented by the other party but counterfactually would have been represented by their own party.

Generally, we will want to evaluate the fairness of a specific districting plan $q$ versus a distribution of counterfactuals.
These counterfactuals could arise from a posited probability distribution capturing the factors relevant to a non-partisan redistricting process, or they could be a set of alternative redistricting plans that were considered as part of the legislative process leading to a final enacted plan.
We adopt the notation $F(q)$ for this scenario: $$
    F(q) \dfeq \E[H\mid V(I) = \dem, Q=q] - \E[H\mid V(I) = \rep, Q=q],
$$ so that we recover $F=\E[F(Q)]$.

[**CM: better to define just** $F(Q)$**, since we will rarely compare two sets of plans?**]

In the districting plan examples of Figure \ref{fig:mini-harm}, the fairness of the Democratic gerrymander of Mininois is $-0.519$, the fairness of the Democratic gerrymander of Minissouri is $-0.415$, and the fairness of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).

Despite being derived from a local harm measure, $F(Q)$ is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
Yet unlike a simple measure of seats, $F(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to partisan fairness.
In other words, $F(Q)$ provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
This connection further supports the use of total seats as a summary statistic in redistricting simulation methods, as advocated by @deford2020symmetry.

```{=tex}
\begin{restatable}{prop}{propfairness} \label{prop:fairness}
Basically, if turnout is equal across districts, for a districting plan $q\in\Plans$ and political party $p\in\Parties$, we have
\begin{align*}
    F(q) = \E[\seats(p, V, Q)\mid Q=q] - \E[\seats(p, V, \tilde Q)].
\end{align*}

\end{restatable}
```
# Calculating the Harm and Fairness Measures {#sec:estimating}

The measures $H$ and $F(q)$ defined above require specification of a probability measure on elections and districting plans, $\pi$.
Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
Either way, a collection of past election data and an enacted districting plan alone are not enough to calculate $H$ and $F(q)$.

While not strictly necessary, modelling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
Practically, this translates to the assumption that individual voters would vote for the same political party no matter the districting plan.
To the extent that voters reward or punish an incumbent party for the districting plan they adopt, or that an adopted districting plan affects which candidates run in a district, this assumption is violated.
But as our suggested model for election outcomes will not account for candidate factors or policy preferences, the marginal effect on our conclusions will be minimal.

## Modelling Districting Plans

1.  discrete and no universal metric structure

2.  algorithms developed to sample from maximum-entropy distributions [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020]

3.  impose constraints required by law and tradition; then samples are exchangeable

## Modelling Election Outcomes

1.  complex, with many unknown future factors

2.  need simple model that will capture primary factor (nationwide political environment).
    Ideally would be easy to implement

3.  we propose **Proportionate Uniform Shift**, where a constant fraction of voters of one party are assumed to switch to the other.

    1.  Not mathematically equivalent Butler's Uniform Partisan Swing [@butler1951swing], but near-identical for most precincts and districts.

    2.  Highly accurate in out-of-sample prediction, considering simplicity [@katz2020].

    3.  Can be viewed as mapping vote shares through a Laplace CDF centered at 0.5 with rate 2

    4.  Model shift amounts as Normal.
        Fit to historical variability in nationwide vote shares to get $\mathrm{t}_{22}(0, 0.03723)$.

4.  More complicated models certainly possible and straightforward (if not particularly efficient) to compute.
    Could e.g. model district deviations as a systematic national shift plus a local idiosyncratic effect, like the Stochastic Uniform Partisan Swing of @gelman1994unified.

5.  Formulas to actually compute these things are in the Appendix:

    1.  Global utility is discrete, so we can compute the range of shifts that produce each number of seats, and the average vote share for those ranges.
        Combine these

    2.  To be computationally efficient, local utility requires linearizing the vote share curve.

    3.  For voter fairness, PUS means that relative share of party's voters in each precinct remains the same.
        Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)

# Connections to Partisan Symmetry and Fairness, Standards {#sec:connections}

Over the past decade [**CM: correct?!**], researchers have developed a veritable zoo of measures to identify partisan gerrymanders. Specifically, these measures aim to quantify deviations from partisan symmetry, fairness, or competitiveness. Some of the most commonly used are perhaps the *efficiency gap* [@stephanopoulos2015], *partisan bias* [@tufte1973pb; @king1987democratic], and the *mean-median* measure [@mcdonald2015unfair; @wang2016three].
All of these measures may be estimated using district vote shares.

## The Efficiency Gap

1.  has been criticized [@tamcho2017egap; @bernstein2017egap; @deford2020symmetry]

## Partisan Bias

1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
2.  Focuses on only one part of S-V curve, which may not be relevant

## Mean-Median

1.  Like PB, focuses on only one part of S-V curve, which may not be relevant

## Districting Plan Competitiveness {#sec:competitive}

The notion of district *competitiveness*, while related to the slope of the seats-vote curve around $\half$, is not the same as partisan symmetry.
However, one commonly used measure of competitiveness, from @tamcho2016, is related to our local utility measure $U^{(L)}$.
The measure of @tamcho2016, which we will denote by $\mathrm{comp}_\alpha(v,q)$ contains two terms: one, $T_p(v,q)$, measuring the closeness of district vote shares to 50%, and the other, $T_e(v,q)$, measuring the closeness of the overall share of seats to 50%; the relative importance of these two terms is controlled by $\alpha$.
It is this first term which his closely related to $U^{(L)}$, as stated in Proposition \ref{prop:comp} below.

```{=tex}
\begin{restatable}{prop}{propcomp} \label{prop:comp}
Basically, if turnout is equal across districts, for an election $v\in\Elecs$ and districting plan $q\in\Plans$ and two political parties, we have
\begin{align*}
    \E[U^{(L)}\mid V=v, Q=q] &= \frac{1}{2|D|}\sum_{d\in D}
        \qty|\frac{\votes^{(p_a)}}{\votes^{(p_b)} + \votes^{(p_b)}}-\half| \\
    &= \half T_p(v,q) = \frac{\mathrm{comp}_\alpha(v,q)}{1+\alpha T_e(v_q)}.
\end{align*}

\end{restatable}
```
Importantly, $\mathrm{comp}_\alpha(v,q)$ is designed to be *minimized*---smaller values correspond to more competitive districts.
But, as briefly discussed above, this actually runs counter to the local utility measure, which encourages lopsided districts with as many similarly-minded voters as possible.
@brunell2010 discusses this point in much more detail, arguing that competitive districts should be avoided where possible.
Buttressing our choice of local utility function, he analyzes ANES survey data to show that voters approve more highly of their representative and of Congress overall when their representative votes in alignment with their preferences.

# Real-world Example {#sec:empirical}

1.  New Jersey

    1.  Not a 50-50 state

    2.  moderate number of districts

```{r nj-vars, fig.cap="Various measures for an ensemble of New Jersey districting plans.", fig.width=6.5, fig.height=6.5}
include_graphics(here("out/nj_vars.pdf"))
```

# Discussion {#sec:discussion}

\appendix

# Proofs of Propositions

```{=tex}
\propglobal*

\begin{proof}
\end{proof}

\propfairness*

\begin{proof}
\end{proof}

\propcomp*

\begin{proof}
\end{proof}
```
